Uncertainty propagation for predictor effects and bias correction in generalized linear (mixed) models

### 2022-01-24 manuscript/vareffects_plosone.tex
varpred manuscript

In many application, outcome predictions or predicted probabilities is of interest. For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated predictor effects are sensitive to how the predictions are generated. The most common approach is generating predictor effects at a ``typical value'', usually the mean, of non-focal predictor, i.e., mean-anchored which is the effect of an ``average case'' in the population. In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-anchored approach generate predictor effects that are biased and not consistent with the observed quantities. An alternative is the population-based approach which estimates the average effect in the population. Moreover, isolated confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that population-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.

@@

In many applications, outcome predictions or probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret and less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other variables.

@@

When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation. Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of \emph{focal} predictor(s) and appropriate \emph{model center} for \emph{non-focal} predictors especially in multivariate models
\item  propagation of uncertainty -- can we incorporate uncertainty in nonlinear components of GL(M)Ms? Should we exclude variation in non-focal parameters?
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their typical values (e.g., averages) \cite{hanmer2013behind}. This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} and \emph{estimated marginal means} \cite{lenth2018package}. In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor. Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.

The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, averages the non-focal predictors. However, there are a number of choices a number of choices which one has to make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}). We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -0.5$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the observed average i.e., $\bar{y}$. Fig~\ref{fig:justify_plots} compares the predictors effects of the two models.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{justify_plots-figure1.pdf}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the mean observed $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors. The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the observed in all the three approaches. B: Even with the simple interaction between the non-focal predictors, we start seeing larger deviation of the predicted mean, $\bar{\hat{y}}$, from the observed, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}

@@

An alternative to averaging the non-focal predictors, is the \emph{population-based} approach which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.

In some applications, we may only be interested in the uncertainties associated with a particular focal predictor -- \emph{isolated} confidence intervals. However, currently, the two packages do not provide straightforward way do achieve this. In addition to prediction lines shown in Fig~\ref{fig:justify_plots}A above, we may want to include the confidence bands associated with the predictions from  all the three methods, as shown in Fig~\ref{fig:justify_ci_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{justify_ci_plots-figure2.pdf}
\caption{{\bf A comparison of regular and isolated confidence bands.} The description of horizontal, vertical and trend lines remain the same as above. The wider dotted blue curves overlaying the green curves are the regular confidence bands from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor are the isolated confidence bands from \pkg{varpred}. For simple models, the isolated confidence bands crosses at the model center.}
\label{fig:justify_ci_plots}
\end{figure}

@@

When dealing with nonlinear link functions, the correct predictions, for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions.

The main purpose of this article is to discuss and implement various approaches for computing predictor effects and provide an alternative method for computing the associated confidence intervals. We further explore approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{qoi_age_pred_plot-figure3.pdf}
\caption{{\bf A comparison of predictor effect and unconditional marginal effect of age on household size.} The black line is the predictor effect trend line line, while the dotted lines indicate the changes along the \code{age} and \code{household size} axis. For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
\label{fig:qoi_age_pred_plot}
\end{figure}

@@

To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in predictor effects of a particular predictor, i.e., focal, $x_f$, from the set of predictors. To keep it simple, assume that the model has no interaction terms. The idea is to \emph{anchor} the values of non-focal predictors to some particular values. For example fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors. We refer to this as mean-anchored approach. The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor(s).

@@

What about the confidence intervals (CI)? The limits of the confidence intervals are points, not mean values. In principle, every value of focal predictor has a different CI. The traditional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$ \cite{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution. This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.  But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$, i.e., isolated confidence intervals?

@@

In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response. However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable. In such cases, bias correction is needed when back-transforming the predictions to the original scales. Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}. Here, we describe and implement a different approach, population-based approach for bias correction.

@@

We start by fitting two models from the simulated data, i.e., consider the polynomial predictor, age, as a focal predictor and also, consider the non-polynomial predictor, wealth index as the focal predictor. After fitting the model, we constructed predictor effects and then compared isolated vs non-isolated confidence intervals associated with these predictions (see Fig~\ref{fig:pred_cubic_plots}).

@@

In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and cross, in the case of isolated confidence bands, at the model center (see Fig~\ref{fig:pred_cubic_plots}B). On the other hand, if the focal predictor is characterized by interactions or any other form of transformation, the model center is not necessarily a point in the predictor space. In this case, the isolated confidence bands will be narrower then the non-isolated but not necessarily crossing (see Fig~\ref{fig:pred_cubic_plots}A).

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
\caption{{\bf Mean-anchored and population-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}. In both cases, age and wealth index effect, the average predicted probability of improved water based on population-based approach is very close to the observed proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion of improved water quality, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The dotted grey points are binned observations -- observed proportions of improved water quality in each bin. We would expect the all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected population-based approach) of the predictions on the response scale, this may not always be case.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
\caption{{\bf Mean-anchored and population-based predictor effects for mediated effect.} These figures compare the mean-anchored and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and mediated, $z \sim x$, models. A: In the absence of mediator variable, both mean-anchored and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias. Similarly, in both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations. However, the population-based approach still closely approximates the average proportion in the simulated data. The simulated data. The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The dotted grep points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches. However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}. Including both $y$ and $x$ dilutes the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).

@@

Our simulations examples majorly focused on simple linear and logistic models due their wide range of usage and application. In addition, these models act as a starting point for building other complex models, including mixed effect models and models with categorical predictors. However, the logic for extension of approaches to more complex models, including other forms of nonlinear link functions is very straightforward. In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework.

@@

From our theoretical, methodological and simulation results, researchers these kind of models should, in the absence of theoretical justification, report predictions based on population-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question. Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

@@

>>>>

In many applications using genralized linear (mixed) models, outcome predictions or predicted probabilities are often of interest. For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable. A general way to summarize these kind of models is through predictor effects, which are also sensitive to which values of non-focal predictors are chosen. The most common approach is generating predictor effects at a ``typical value'', usually the mean, of non-focal predictor, i.e., mean-anchored which is the effect of an ``average case'' in the population. In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-anchored approach generate predictor effects that are biased and not consistent with the observed quantities. An alternative is the population-based approach which estimates the average effect in the population. Moreover, isolated confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that population-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction.

@@

In many applications using simple or generalized linear (mixed) models, outcome predictions or probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret and less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.

@@

When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation. Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of \emph{focal} predictor(s) and appropriate \emph{anchor} for non-focal predictors especially in multivariate models
\item  propagation of uncertainty -- can we incorporate uncertainty in nonlinear components of GL(M)Ms? Should we exclude variation in non-focal parameters?
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their typical values (e.g., averages) \cite{hanmer2013behind}. This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} and \emph{estimated marginal means} \cite{lenth2018package}. In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor, for an ``average case''. Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.

The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, averages the non-focal predictors. However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}). We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{justify_plots-figure1.pdf}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the mean observed $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors. The grey points are the binned simulated $y$. The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the observed in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. B: Even with the simple interaction between the non-focal predictors, we start seeing deviation of the predicted mean, $\bar{\hat{y}}$, from the observed, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}

@@

In some applications, we may only be interested in the uncertainties associated with a particular focal predictor -- \emph{isolated} confidence intervals. However, currently, the two packages do not provide straightforward way do achieve this. To illustrate this, we generate predictor effect of $x_2$ from model~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence bands, as shown in Fig~\ref{fig:justify_ci_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{justify_ci_plots-figure2.pdf}
\caption{{\bf A comparison of traditional and isolated $95\%$ confidence bands.} The description of horizontal, vertical and trend lines remain the same as above. The wider dotted blue curves overlaying the green curves are the traditional confidence bands from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal predictor, i.e., the model center, are the isolated confidence bands from \pkg{varpred}. For simple models, the isolated confidence bands crosses at the model center.}
\label{fig:justify_ci_plots}
\end{figure}

@@

When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions. An alternative to averaging the non-focal predictors, is the \emph{population-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.

The main purpose of this article is to discuss and implement various approaches for computing predictor effects and provide an alternative method for computing the associated confidence intervals. We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{qoi_age_pred_plot-figure3.pdf}
\caption{{\bf A comparison of predictor effect and unconditional marginal effect of age on household size.} The black line is the predictor effect trend line, while the dotted lines indicate the changes along the \code{age} and \code{household size} axis. For a linear model with no interaction, the marginal effect is the slope of the predictor effect line.}
\label{fig:qoi_age_pred_plot}
\end{figure}

@@

To get an intuition of how conditioning on the mean values of the non-focal predictors work, suppose we are interested in predictor effects of a particular predictor, i.e., focal, $x_f$, from the set of predictors. To keep it simple, assume that the model has no interaction terms. The idea is to \emph{anchor} the values of non-focal predictors to some particular values. For example, fixing the values of non-focal predictor(s) at some typical values -- typically determined by averaging (for now) in some meaningful way, for example, arithmetic mean  for continuous and average over the levels of the factors for categorical non-focal predictors. We refer to this as \emph{mean-anchored} approach. The most convenient way to achieve this is by constructing $\bX^\star$ by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.

@@

What about the confidence intervals (CI)? The limits of the confidence intervals are points, not mean values. In principle, every value of focal predictor has a different CI. The traditional way to compute variances for predictions is $\sigma^2 = \textrm{Diag}(\bX^\star \Sigma \bX^{\star\top})$ \cite{lenth2018package, fox2009effect}, so that the confidence intervals are $\eta \pm q\sigma$, where $q$ is an appropriate quantile of Normal or t distribution. This approach incorporates all the uncertainties -- including the uncertainty due to non-focal predictors.  But what if we are only interested in the uncertainty as a result of the focal predictor, so that the confidence intervals are $\eta \pm q \sigma_f$? We call this isolated confidence intervals.

@@

In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response. However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable. In such cases, bias correction is needed when back-transforming the predictions to the original scales. Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}. Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions. Here, we describe and implement a different approach, population-based approach for bias correction.

@@

We start by fitting two models from the simulated data, i.e., consider a cubic polynomial predictor, age, as a focal predictor and also, consider the non-polynomial predictor, wealth index as the focal predictor. After fitting the model, we constructed predictor effects and then compared isolated vs non-isolated confidence intervals associated with these predictions (see Fig~\ref{fig:pred_cubic_plots}).

@@

In the absence of complex higher order interaction terms or transformations on the focal predictor in simple linear models, we would expect, at the model center, the observed average, $\over{hh~size}$, and the average of the predictions, $\over{\widehat{hh~size}}$, to be identical and cross, in the case of isolated confidence bands, at the model center (see Fig~\ref{fig:pred_cubic_plots}B). On the other hand, if the focal predictor is characterized by interactions or any other form of transformation, the model center is not necessarily a point in the predictor space. In this case, the isolated confidence bands will be narrower than the non-isolated but not necessarily crossing at the model center (see Fig~\ref{fig:pred_cubic_plots}A).

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
\caption{{\bf Mean-anchored and population-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}. In both cases, age and wealth index effect, the average predicted probability of improved water based on population-based approach is very close to the observed proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion of improved water quality, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin. We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected population-based approach) of the predictions on the response scale, this may not always be the case.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
\caption{{\bf Mean-anchored and population-based predictor effects for mediated effect.} These figures compare the mean-anchored and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models. A: In the absence of mediator variable, both mean-anchored and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias. Consequently, in both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations. However, the population-based approach still closely approximates the average proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-anchored, population-based predicted probability and observed proportion, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches. However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}. Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).

@@

Our simulations examples majorly focused on simple linear and logistic models due their wide range of usage and application. In addition, these models act as a starting point for building other complex models, including mixed effect models and models with categorical predictors. However, the logic for extension of approaches to more complex models, including other forms of nonlinear link functions is very straightforward. In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.

@@

From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on population-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question. Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

@@

\section*{Acknowledgments}

This work was supported by a grant to Jonathan Dushoff from the Natural Sciences and Engineering Research Council of Canada (NSERC) Discovery.

\section*{Author Contributions}

\textbf{Conceptualization:} Jonathan Dushoff

\noindent\textbf{Software:} Steve Cygu, Benjamin M. Bolker

\noindent\textbf{Supervision:} Jonathan Dushoff, Benjamin M. Bolker

\noindent\textbf{Writing – original draft:} Steve Cygu
